---
title: "BRS prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BRS prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!--	To manually compile this RMarkdown document: -->
<!--	devtools::build_vignettes() -->

# Introduction

This R Markdown document is a step-by-step to reproducing the results from the original publication, as well as showing key gene expression results, handling pamr-object input/output, and generalizing the method to new datasets.

## Citation

If you use ```BRSpred```, please use its corresponding citation at:
_de Jong, et al. Submitted_

# BRS pipeline

## Installation

```{r github}
# devtools::install_github("CostelloLab/BRSpred")
```

## Setup

```{r setup}
library(BRSpred)
```

## Datasets

The package comes with two main dasets:
* Cohort A (training) from Erasmus MC (N=132 total)
* Cohort B (testing) from Erasmus MC (N=151 total)


```{r}
data(CohortA.vst)
data(CohortB.vst)
```

The clinical information for the patients from Erasmus MC are stored in:
```{r clinical}
data(erasmus.clinical)
table(erasmus.clinical$Cohort)
```

## Training

Select top 2000 genes based on their variance:
```{r}
topVarGenes <- head(order(matrixStats::rowVars(as.matrix(CohortA.vst)), decreasing = TRUE), 2000)
TopVar.A  <- CohortA.vst[topVarGenes, ]
TopVar.A2 <- t(scale(t(TopVar.A)))
```

Use unsupervised consensus clustering on the training cohort (A) to discover the subgroups:

```{r, fig.fullwidth = TRUE, fig.width=7, fig.height=7}
consensus.res <- ConsensusClusterPlus::ConsensusClusterPlus(TopVar.A2, maxK=5, reps=2500, pItem=.95, pFeature=.95, clusterAlg="pam", distance="pearson", seed=123)
icl <- ConsensusClusterPlus::calcICL(consensus.res)
icl[["clusterConsensus"]]
```

Append consensus classes to clinical information and reassign BRS-label numbers to be more intuitive:
```{r}
consensus.df <- as.data.frame(consensus.res[[3]][["consensusClass"]])
Clinicaldata.A.pre <- erasmus.clinical[which(erasmus.clinical$Cohort == "A" & erasmus.clinical$Tumor == "Primary"),]
colnames(consensus.df)[1] <- paste("Erasmus.BRS")
Clinicaldata.A.pre$Erasmus.BRS <- consensus.df$Erasmus.BRS
Clinicaldata.A.pre$Erasmus.BRS <- gsub("^3", "BRS1", gsub("^2", "BRS3", gsub("^1", "BRS2", Clinicaldata.A.pre$Erasmus.BRS)))
Clinicaldata.A.pre$Erasmus.BRS <- factor(Clinicaldata.A.pre$Erasmus.BRS, levels=c("BRS1", "BRS2", "BRS3"))
table(Clinicaldata.A.pre$Erasmus.BRS)
```

Kaplan-Meier curves in the training data:
```{r, message=FALSE, fig.fullwidth = TRUE, fig.width=7, fig.height=7}
fit1 <- survival::survfit(survival::Surv(Time_to_prog_or_FUend, Progression) ~ Erasmus.BRS, data = Clinicaldata.A.pre)
survminer::ggsurvplot(fit1,   
	data = Clinicaldata.A.pre,   
	censor.shape="|",   
	censor.size = 3,  
	xlab = "Time in months",   
	title = "PFS in Cohort A",  
	font.title    = c(16, "bold"),
	font.y = "bold",  
	ylab = "Progression free proportion",  size = 1.2,   conf.int = FALSE,   xlim = c(0,120),
	pval = T, risk.table = TRUE, risk.table.col = "black",
	ggtheme = ggplot2::theme_classic(),
	risk.table.y.text.col = T, risk.table.y.text = FALSE,
	break.time.by = 24, palette = c("forestgreen", "dodgerblue3",  "firebrick")) 
```

## Testing set

Normalize the expression in validation/testing dataset using quantile normalization:

```{r}
# Intersect common genes
CohortA.vst2 <- CohortA.vst[intersect(rownames(CohortA.vst),rownames(CohortB.vst)), ] 
CohortB.vst2 <- CohortB.vst[rownames(CohortB.vst) %in% rownames(CohortA.vst2), ]
# Order genes in Cohort B to be same as in Cohort A
CohortB.vst2 <- CohortB.vst2[match(rownames(CohortA.vst2), rownames(CohortB.vst2)),]
# Quantile normalization
CohortB.vst3 <- preprocessCore::normalize.quantiles.use.target(x = as.matrix(CohortB.vst2), target = c(unlist(CohortA.vst2)))

rownames(CohortB.vst3) = rownames(CohortB.vst2)
colnames(CohortB.vst3) = colnames(CohortB.vst2)

# Expression value quantiles compared to training data prior and post normalization
quantile(as.matrix(CohortA.vst2))
quantile(as.matrix(CohortB.vst2))
quantile(as.matrix(CohortB.vst3))
```

Use `pamr` R-package after matching cohort B to the original top 2000 varying genes:

```{r}
Training_set <- TopVar.A[which(rownames(TopVar.A) %in% rownames(CohortB.vst3)), ]
Testing_B <- CohortB.vst3[ rownames(CohortB.vst3) %in% rownames(Training_set), ]
# Reorder genes to match original 2000 gene ordering
Testing_B <- Testing_B[match(rownames(Training_set), rownames(Testing_B)),]

## Z-score scale data for PAMR
Training_set2 <- t(scale(t(Training_set)))
Testing_B2 <- t(scale(t(Testing_B)))
```

Train the `pamr` supervised model to the discovered labels and plot cross-validation as a function of penalization threshold:
```{r, fig.fullwidth = TRUE, fig.width=7, fig.height=7}
set.seed(1234)
CohortB.train.list <- list(x=as.matrix(Training_set2), y=as.factor(Clinicaldata.A.pre[,"Erasmus.BRS"]), genenames=row.names(Training_set2), samplelabels=colnames(Training_set2), geneids=row.names(Training_set2)) 
pamr3cl <- pamr::pamr.train(data=CohortB.train.list)
mycv <- pamr::pamr.cv(pamr3cl,data=CohortB.train.list)
pamr::pamr.plotcv(mycv)
```

Select optimal threshold based on cross-validation minima, contributing genes and their results in CV:
```{r}
threshold.optimal <- mycv$threshold[max(which(mycv$error == min(mycv$error)))]
threshold.optimal
# Using BRS-package's re-implementation of pamr.listgenes for convenience; no forceful printing to stdout
topgenes <- BRSpred::BRS_pamr.listgenes(pamr3cl, data=CohortB.train.list,threshold=threshold.optimal, fitcv=mycv)
head(topgenes)
```

Performing prediction for cohort B:

```{r}
CohortB_pred <- pamr::pamr.predict(fit=pamr3cl, newx=Testing_B2, threshold=threshold.optimal)
table(CohortB_pred)
```

Kaplan-Meyer using the predictions for Cohort B:
```{r, message=FALSE, fig.fullwidth = TRUE, fig.width=7, fig.height=7}
Clinicaldata.B <- BRSpred::erasmus.clinical[which(BRSpred::erasmus.clinical$Cohort == "B"),]
Clinicaldata.B$CohortB_predict <- CohortB_pred
fit2 <- survival::survfit(survival::Surv(Time_to_prog_or_FUend, Progression) ~ CohortB_predict, data = Clinicaldata.B)
survminer::ggsurvplot(fit2,
	data = Clinicaldata.B,
	censor.shape="|", censor.size = 3,
	xlab = "Time in months", title = "PFS in Cohort B", ylab = "Progression free proportion",
	font.title = c(16, "bold"),
	font.y = "bold", size = 1.2, conf.int = FALSE, xlim = c(0,120),
	pval = T, risk.table = TRUE, risk.table.col = "black", 
	ggtheme = ggplot2::theme_classic(),
	risk.table.y.text.col = T, risk.table.y.text = FALSE,
	break.time.by = 24, palette = c("forestgreen", "dodgerblue3", "firebrick")) 
```

## Heatmaps

We predefine some subsets of genes of interest and plot their expression in both the training and validation cohort:

```{r}
genesets <- list(
	"EMT" = c("CDH3", "CLDN3", "CLDN4","CLDN7", "TWIST1", "SNAI1", "VIM", "ZEB1", "ZEB2"),
	"Basal" = c("CD44", "KRT5", "KRT6A", "KRT14","KRT16", "DDR2"),
	"Immune" = c("CD4","CD8A","CD19", "CD79A","CD80", "CD86", "CD274","CTLA4", "PDCD1","FOXP3","TGFB1","TNF","IDO1","NFKB1", "JAK3", "CSF1R"),
	"HLA" = c("HLA-A","HLA-B","HLA-C","HLA-DPA1","HLA-DPB1","HLA-DOA","HLA-DQA1", "HLA-DRB1"),
	"Luminal" = c("PPARG", "FOXA1", "GATA3", "ERBB2", "ERBB3","DDR1", "UPK1A", "UPK2","UPK3B", "UPK3A", "KRT18", "KRT19","KRT20", "CDH1"),
	"FGFR3" = "FGFR3",
	"MYC signaling" = c("MYC", "MYCL", "MYCN", "DUSP2", "EXOSC5", "PPAN", "MYBBP1A", "HK2", "NOC4L", "PUS1", "FARSA"),
	"Cell Cycle" = c("PLK1","E2F1","CDK1","CDK2","CDC7","CCNB1", "CCNB2","CCNE1","E2F3", "EZH2","CCNE2", "ORC5", "RB1","RFC3","STAG2", "BUB1"),
	"Autophagy" = c("MAP1LC3B","ATG3","ATG4C", "ATG5", "ATG7","ATG12","BECN1","MTOR","PIK3CA","PIK3C3"),
	"Intracellular transport" = c("RAB5A", "SEC22B", "VAMP4", "VAMP7", "SCAMP1", "ADAM10", "STX7", "STX12", "TOM1L1", "TMX1", "TSPAN8", "BRAF"),
	"Antigen processing" = c("GLMN","UBE2K","UBE2W","FBXO30","NEDD4L","PSMA1","UBA5","UBA6","FBXL5", "LTN1", "RNF182","RNF138")
)
```

### Cohort A heatmap

Construct geneset-average expression within samples, based on the trained clustering classes (we order samples based on BRS-classes and inside each class based on progression status):
```{r}
suborderA <- rownames(Clinicaldata.A.pre[order(Clinicaldata.A.pre$Erasmus.BRS, Clinicaldata.A.pre$Progression),])

exprsmatA <- do.call("rbind", lapply(genesets, FUN=function(genes){
	t(as.matrix(colMeans(CohortA.vst[genes,suborderA], na.rm=TRUE)))
}))
rownames(exprsmatA) <- names(genesets)
```

These are then heatmap-plotted e.g. using ```pheatmap```-package, showing clear trends connected to the BRS-classes:

```{r, message=FALSE, fig.fullwidth = TRUE, fig.width=7, fig.height=7}
breaksList = seq(-1, 1, by = 0.2)
metadata<-Clinicaldata.A.pre[,c("Progression","Erasmus.BRS")]
pcols = list( 
	Progression = c("No Progression"="limegreen", "Progression"="red"),
	Erasmus.BRS = c("BRS1"="forestgreen","BRS2"="dodgerblue3", "BRS3"="firebrick") 
)

pheatmap::pheatmap(exprsmatA, cluster_rows = FALSE, cluster_cols = FALSE, scale = "row",
         show_colnames = FALSE,
         show_rownames = TRUE,
         fontsize = 8,
         color = colorRampPalette((RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")))(length(breaksList)),
         breaks = breaksList,
         clustering_distance_rows = "euclidean",
         color.scheme.symmetric = T,
         gaps_row = 1:10,
         gaps_col = which(!duplicated(metadata[suborderA,"Erasmus.BRS"]))[-1]-1,
         annotation_col = metadata,
         annotation_colors = pcols)        
```

### Cohort B heatmap

We first construct geneset-average expression within samples, with ordering based on predicted BRS classes:

```{r}
suborderB <- rownames(Clinicaldata.B[order(Clinicaldata.B$CohortB_predict, Clinicaldata.B$Progression, decreasing = FALSE), ])

exprsmatB <- do.call("rbind", lapply(genesets, FUN=function(genes){
	t(as.matrix(colMeans(CohortB.vst[genes,suborderB], na.rm=TRUE)))
}))
rownames(exprsmatB) <- names(genesets)
```

These are then heatmap-plotted e.g. using ```pheatmap```-package, showing clear trends connected to the BRS-classes:

```{r, message=FALSE, fig.fullwidth = TRUE, fig.width=7, fig.height=7}
metadata<-Clinicaldata.B[,c("Progression","CohortB_predict")]
pcols = list( 
	Progression = c("No Progression"="limegreen", "Progression"="red"),
	CohortB_predict = c("BRS1"="forestgreen","BRS2"="dodgerblue3", "BRS3"="firebrick") 
)

pheatmap::pheatmap(exprsmatB, cluster_rows = FALSE, cluster_cols = FALSE, scale = "row",
         show_colnames = FALSE,
         show_rownames = TRUE,
         fontsize = 8,
         color = colorRampPalette((RColorBrewer::brewer.pal(n = 9, name = "YlOrRd")))(length(breaksList)),
         breaks = breaksList,
         clustering_distance_rows = "euclidean",
         color.scheme.symmetric = T,
         gaps_row = 1:10,
         gaps_col = which(!duplicated(metadata[suborderB,"CohortB_predict"]))[-1]-1,
         annotation_col = metadata,
         annotation_colors = pcols)         
```

# BRS for independent datasets

The main function for performing pre-trained BRS prediction for external datasets is the ```BRS```-function. Type ```?BRS``` in R to find out more about the syntax of this function, but essentially the only input required is an appropriately normalized gene expression matrix with gene symbols as rows and samples as columns.

## Missing values and lack of gene overlap

If the ```BRS``` prediction wrapper function cannot find gene symbols used in the original ```pamr``` clustering, it will impute median values for these genes from the original training gene expression matrix. The wrapper will throw a warning, but not an error; a low overlap in identified genes (e.g. lack of expression in some key genes) may therefore result in poor prediction performance, so these warning messages ought to be investigated accordingly for any given cohort.

# Session info

```{r}
utils::sessionInfo()
```
